\section{Conclusion}
\label{sec:conclusion}

We have presented \textsc{predictor4}, a layered decision procedure for equational
implication over magmas that replaces two unsound or incomplete components of its
predecessor with theoretically grounded alternatives.

The central theoretical contribution is the \emph{soundness invariant} for
Knuth-Bendix completion: every rewrite rule $\ell \to r$ generated during the
completion process must satisfy $\operatorname{vars}(r) \subseteq \operatorname{vars}(\ell)$.  We proved
(Lemma~\ref{thm:collapse} and Theorem~\ref{thm:invariant}) that violation of this
invariant causes complete normal-form collapse, making the equality test trivially
true for all equation pairs and thereby producing an unlimited supply of false
positives.  Enforcing the invariant as a guard---rejecting any invalid rule and
returning failure for the affected case---restores the classical soundness guarantee
of KB completion (Theorem~\ref{thm:kb-soundness}).

For the negative path, we established that Z3 SMT model finding is complete for
each fixed domain size $n$ (Theorem~\ref{thm:z3-complete}): it either finds a
size-$n$ counterexample or proves none exists, providing a stronger guarantee than
heuristic hill-climbing.

Empirical evaluation confirms that \textsc{predictor4} achieves $100\%$ accuracy
with zero false positives on stratified samples of 50 and 200 cases from the ETP
ground truth.  The KB component contributes formally verified proofs for approximately
$12\%$ of implication cases after the invariant fix, while the remaining cases are
handled by the BFS fallback or downstream negative tests.

\para{Future work.}
Increasing KB convergence rates (through longer timeouts, alternative reduction
orders, or equivalence-class caching), extending Z3 to larger domain sizes, and
exporting KB proofs as Lean 4 certificates compatible with the ETP formalization
are the most promising directions for future development.  On the theoretical side,
characterizing the class of single-equation magma systems for which KB completion
terminates remains an open problem with connections to the wider theory of term
rewriting decidability.
